### 2021-02-12 manuscript.Rmd
initial version of OCRbayes

---
output: word_document
bibliography: /home/xiang/Documents/Bibliography/seahorse.bib
---

Bayesian hierarchical modeling for Seahorse extracellular flux oxygen consumption rate data analysis

@@

Mitochondrial dysfunction is involved in many complex diseases. 
Efficient and accurate evaluation of mitochondrial functionality is crucial for understanding pathology as well as facilitating novel therapeutic developments. 
As a reliable platform, Seahorse extracellular flux (XF) analyzer is widely used for measuring mitochondrial oxygen consumption rate (OCR) in living cells. 
A hidden feature of Seahorse XF OCR data is that it has a complex data structure, caused by nesting and crossing between multiple factors. 
Surprisingly, statistical analysis of Seahorse XF data has not received sufficient attention, and current methods completely ignore the complex data structure, impairing the robustness of statistical inference. 
To rigorously incorporate the complex structure into data analysis, here we developed a Bayesian hierarchical modeling framework and demonstrated its applicability for analyzing Seahorse XF OCR data with two in-house experimental data sets. 
We showed that our approach can propagate uncertainty in the OCR measurements into the posterior estimation of bioenergetic measures, and is flexible to be applied to analyze experimental data with various experimental designs and sample sizes.

@@

Mitochondria produce energy primarily by transferring electrons along the electron transport chain (ETC) on the inner membrane. 
Along the ETC, there are four complexes (including complex I, II, III and IV), building up a proton gradient that is ultimately used by ATP synthase to generate ATP. 
Since electrons are primarily accepted by O~2~ to produce H~2~O, the activity of ETC can be assessed by measuring oxygen consumption rate (OCR). 
As a reliable and efficient platform, Seahorse XF analyzer is a multiwell plate based respirometry assay that is widely used to quantify OCR in living cells [@Wu2007]. 
Typically, the seahorse XF analyzer measures the OCR of cells in a 96-well plate under different ETC perturbation scenarios, used for assessing mitochondrial functionality such as maximal respiration, leak respiration and ATP-linked respiration. 
Seahorse XF Cell Mito Stress Test consists of four phases including an initial phase, followed by three consecutive injections, including oligomycin, FCCP (carbonyl cyanide-4-(trifluoromethoxy)phenylhydrazone) and a mixture of rotenone and antimycin A. 
In the initial phase, no compound is injected and ETC is not perturbed. 
Once oligomycin is injected in the assay, proton translocation through ATP synthase is blocked, leading to a reduction in OCR. 
Next, injection of FCCP allows protons to move into the mitochondrial matrix independent of the ATP synthase, pushing the oxygen consumption to the maximum level. 
In the end, a mixture of rotenone and antimycin A is injected to inhibit complex I and III, shutting down mitochondrial respiration.

A standard Seahorse run includes three measurement cycles for each phase. 
Every measurement cycle starts by lowering the cartridge and creating a temporary semi-closed 7 $\mu$L chamber [@Gerencser2009]. 
During a measurement cycle, fluorescent oxygen sensors capture oxygen concentration changes in the chamber and output OCR. 
The Seahorse XF analyzer measures OCR in tens to hundreds of thousands of cells per well and typically requires 4–5 replicate wells per experimental group [@Artyomov2020].
A single 96-well plate can contain multiple experimental groups. 
Since measurements are nested within phases, phases are crossed with wells, and wells are nested in experimental groups, a very complex structure is inherently embedded in the Seahorse XF OCR data. 
Surprisingly, current Seahorse XF data analyses most often ignore this complex structure, and by default, data sets are often chopped into subgroups followed by performing ANOVA-like statistical tests. 
Other Seahorse XF data analysis tools such as OCR-Stats [@Yepez2018], SHORE [@Nicholas2017] and Bioenergetic Health Index [@Chacko2014] also do not take the complexity of the data structure into account.

A direct and rigorous way to incorporate the complex structure into data analysis is to use hierarchical modeling. 
Hierarchical models have been extensively developed for analyzing gene expression data  [@Robinson2010a;@Love2014a]. 
Here we developed a Bayesian hierarchical modeling framework for the Seahorse XF OCR data. 
Our modeling framework is based on a fully Bayesian approach due to its flexibility and convenience for constructing the hierarchical models. 
We used two in-house research cases to demonstrate the applicability of our approach analyzing Seahorse XF OCR data.

@@

## Cell culture

Fibroblasts from controls and patients etc.

## Searhorse extracellular flux analysis

Cells were plated the day before the assay in 96 well plates etc…

## Cell number quantification

In both example studies, after Seahorse assays, cells in plate were stained with 4 $\mu$M of 2'-(4-ethoxyphenyl)-5-(4-methyl-1-piperazinyl)-2,5'-bi-1H-benzimidazole trihydrochloride trihydrate (Hoechst 33342, #B2261, Sigma-Aldrich) dissolved in assay medium, for 1 hr at 37°C non-CO 2 incubator. 
Images were taken by Cytation 1 Imaging Reader (BioTek) and cell number was determined by ImageJ 1.52.

## Case study 1

In the first study, fibroblast cells derived from a healthy subject were cultured in galactose exposure medium. 
A SIRT5 inhibitor, DK104e, or the control counterpart, DK104e[O] was administrated into the cell culture. 
Fibroblast treated with DMSO was used as the control. 
In this experiment, four dosages (10 $\mu$M, 20 $\mu$M, 60 $\mu$M and 70 $\mu$M) were used. 
Respiration of fibroblasts was measured using a Seahorse XFe96 Extracellular Flux Analyzer and a Mito Stress test was performed by following the manufacturer’s protocols.

During the quality control, we removed three wells (2 derived from 10 $\mu$M DK104e[O] group and 1 from DMSO group) due to negative OCR values. 
After quality control, the number of wells that contain cells treated by 10 $\mu$M, 20 $\mu$M, 60 $\mu$M and 70 $\mu$M DK104e were 10, 10, 10 and 9, respectively. 
Regarding DK104e[O], the number of wells that correspond to 10 $\mu$M, 20 $\mu$M, 60 $\mu$M and 70 $\mu$M were 8, 10, 9 and 9. 
There are 11 wells in which fibroblasts were treated with DMSO.

## Case study 2

In the second study, fibroblast cells derived from four healthy subjects and two patients with two homozygous loss-of-function variants in *SIRT5* were used for Seahorse XF analysis. 
One patient has a missense variant (c.340C>A, NM_012241.3) which leads to a proline114 substitution to threonine (P114T) in the SIRT5 protein. 
The other patient has missense mutation (c.382C>G, NM_012241.3) resulting in leucine128 to valine change (L128V). 
In this experiment, two injections were used. 
In particular, 1 $\mu$M oligomycin (#O4876, Sigma-Aldrich) and a mixture of 2.5 $\mu$M antimycin (#A8674, Sigma-Aldrich) and 1.25 $\mu$M rotenone (#R8875, Sigma-Aldrich).

For every fibroblast cell line derived from healthy controls, 8 replicate wells were used for galactose exposure and 7 replicate wells were used for glucose exposure. 
For each patient-derived fibroblast cell line, 8 replicate wells were used for galactose and glucose exposure, respectively.

@@

The Bayesian hierarchical models were implemented in Stan (version 2.19.3) [@Carpenter2017]. 
We fitted the model by running Hamiltonian Markov Chain Monte Carlo. 
We ran four Markov chains with 2000 iterations in each chain. 
Results were presented with the posterior mean with 95% credible interval. 
The code can be found at https://github.com/XiangZhangSC/seahorse

@@

$$\text{Basal respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, rotenone}$$
$$\text{ATP-linked respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, oligomycin}$$
$$\text{Proton leak} = \text{OCR}_\text{per 1k cells, oligomycin} - \text{OCR}_\text{per 1k cells, rotenone}$$
$$\text{Spare respiratory capacity} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, initial}$$
$$\text{Maximal respiration} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, rotenone}$$

@@

## Bayesian hierarchical modeling framework for Seahorse OCR data analysis

In this study, we developed a Bayesian hierarchical modeling framework for the Seahorse XF OCR data. 
We applied a fully Bayesian approach due to its flexibility and convenience for constructing the hierarchical models. 
The mathematical description of our Bayesian hierarchical modeling framework is shown below:

$$ \text{OCR}_\text{obs}\text{[i]} \sim \text{Lognormal( logOCR}_\text{true}\text{[W[i],I[i]]}, \text{logOCR}_\text{se}\text{[W[i],I[i]])}$$

$$\text{exp(logOCR}_\text{true}\text{[W[i],I[i]])} = \text{OCR}_\text{per 1k cells}\text{[G[i], I[i]]} \times \text{N}_\text{cells}\text{[W[i]]}$$

$$\text{OCR}_\text{per 1k cells}\text{[G[i], I[i]]} \sim \text{Normal}(\mu\text{[i]}, \sigma\text{)} $$

$$\mu\text{[i]} = \text{Regression model depending on the experimental design}$$

$$\sigma \sim \text{Exponential(1)}$$

$$\text{logOCR}_\text{se}\text{[i]} \sim \text{prior distribution}$$

In Equation 6, we used a lognormal distribution with the true OCR value at the log scale ($\text{logOCR}_\text{true}$) and the measurement uncertainty (also at the log scale, $\text{logOCR}_\text{se}$) to model each observed OCR value ($\text{OCR}_\text{obs}$), because OCR values must be positive. 
$\text{OCR}_\text{obs}$ is a vector of length $\text{N}_\text{wells} \times \text{N}_\text{injections} \times \text{N}_\text{Measurement}$, where $\text{N}_\text{wells}$, $\text{N}_\text{injections}$ and $\text{N}_\text{Measurement}$ are number of wells, injections and measurement cycles, respectively. 
The $\text{logOCR}_\text{true}$ and $\text{logOCR}_\text{se}$ are matrices of $\text{N}_\text{wells}$ rows and $\text{N}_\text{injections}$ columns.

In Equation 7, we modeled the true OCR value as a function of the cell number. 
$\text{N}_\text{cells}$ is a vector of length $\text{N}_\text{wells}$, and every entry represents the cell number in that well. 
$\text{OCR}_\text{per 1k cells}$ is a matrix with $\text{N}_\text{Groups}$ rows and $\text{N}_\text{injections}$ columns. 
Every entry of this matrix represents OCR value per 1000 cells that received the same treatment (or from the same group) and injection.

In Equation 8 and 9, we applied a linear regression model in which the OCR value per 1000 cells was used as the response (or dependent) variable to estimate effect sizes of various factors (e.g. treatment, group). 
Since experimental design and plate set-up are not uniform between studies, regression models need to be constructed accordingly. 
The parameter $\sigma$ represented the residual that cannot be explained by the regression model. 
We used a generic the prior distribution ($\text{Exponential(1)}$) for $\sigma$.

In this study, since our case studies focused on human fibroblast cells, we used a informative prior distribution for $\text{logOCR}_\text{se}$. 
We estimated the prior distribution based on the Seahorse XF OCR data in OCR-stats [@Yepez2018] by running the fitdistr function built in the MASS r package. 
As a result, the $\text{Lognormal(-3.11, 0.81)}$ was used as the prior distribution for $\text{logOCR}_\text{se}$.

## Case studies

To demonstrate how our Bayesian hierarchical modeling approach can be used for Seahorse XF OCR data analysis, we used two in-house generated data sets that investigate mitochondrial function alteration due to either treatment of a novel inhibitor targeting SIRT5 (case 1) or a loss-of-function mutation in *SIRT5*.

### Case study 1: Factorial design with no biological replicates

In the first case study, we assessed if the SIRT5 inhibitor DK104e could lead to mitochondrial dysfunction in human fibroblasts.
This study used a factorial design with two factors: treatment and injection. 
In addition, a dosage escalation was applied in this study. 
Here, we treated the dosage factor as a continuous variable. 
Based on the experimental design of this study, we composed the regression model as:

$$\mu\text{[i]} = \beta_0 + \beta_T\text{[T[i]]} + \beta_D \text{D[i]} + \beta_I\text{[I[i]]}$$

$\beta_0$ is the overall average of OCR per 1000 cells in all wells and across injections. 
Based on the OCR-stats data, we estimated the prior distribution for $\beta_0$ as $\text{Lognormal(0.54, 0.32)}$. 
$\beta_T$ is the deviation in OCR due to treatment, DK104e (SIRT5 inhibitor) or DK104e[O] ("placebo"). 
$\beta_D$ is the effect when the dosage increases 10$\mu$M. 
We assume that the higher dosage the larger effect will be. 
$\beta_I$ is the deviation in OCR due to injection, including initial, oligomycin, FCCP and rotenone/antimycin A.

Our analysis showed that in DMSO-treated fibroblasts, the median basal respiration, ATP-linked respiration, proton leak, spare respiratory capacity and maximal respiration were 3.23 [95% credible interval 30.09, 3.34] pmol/min, 3.14 [2.99, 3.25] pmol/min, 0.0955 [0.0258, 0.158] pmol/min, -0.890 [-1.05, -0.705] pmol/min, and 2.34 [2.22, 2.48] pmol/min, respectively. 
In general, treating fibroblasts with 10-60 $\mu$M SIRT5 inhibitor Dk104e or "placebo" Dk104e[O] resulted in similar basal respiration, ATP-linked respiration, proton leak, spare respiratory capacity and maximal respiration. 
We found that the proton leak in 10 $\mu$M Dk104e treated fibroblasts was 0.114 [0.0143, 0.203] pmol/min, whereas in the "placebo" group (10 $\mu$M Dk104[O]) that value was 0.368 [0.101, 0.469] pmol/min. 
However, the difference was not statistically significant (mean difference -0.255 [-0.387, 0.0185]) (Figure 1). 
In contrast, we found that at 60 $\mu$M the proton leak in Dk104e treated fibroblasts was 0.144 [0.113, 0.181] pmol/min, 0.0736 [0.00898, 0.143] pmol/min higher than the proton leak (0.0711 [0.0124, 0.128]) in Dk104e[O] treated fibroblasts (Figure 1). 
At 70 $\mu$M, Dk104e treated fibroblasts showed increasing in basal respiration (1.45 [1.19, 1.68] pmol/min), ATP-linked respiration (0.793 [0.537, 1.08]), proton leak (0.662 [0.446, 0.784] pmol/min) and maximal respiration (1.09 [0.775, 1.40] pmol/min) (Figure 1).

![Figure 1 Posterior distributions of basal respiration, ATP-linked respiration,  proton leak, spare respiratory capacity and maximal respiration in galactose-cultured fibroblasts treated with either SIRT5 inhibitor Dk104e or Dk104e$\text{[O]}$](./inhibitor_study_respiration.jpg)

## Case study 2: Split-plot design

The second study was an experiment in which fibroblasts derived from two patients with *SIRT5* mutations were compared to fibroblasts derived from four healthy individuals. 
The goal of this study was to identify mitochondrial dysfunction in the *SIRT5* mutation carriers. 
In this study, subjects were nested in group factor (healthy control or patient). 
Meanwhile, treatment factor (galactose and glucose) and injection factor were crossed with subjects. 
With both nested and crossed factors, this experiment essentially has a split-plot design. 
The split plot design was originally invented in agricultural experiments in order to deal randomization of hard-to-change factors. 
Accordingly, we specified the regression model as:

$$\mu\text{[i]} = \beta_0 + 
\beta_G\text{[G[i]]} + 
\beta_T\text{[T[i]]} + 
\beta_I\text{[I[i]]} + 
\beta_{IT}\text{[I[i],T[i]]} + 
\beta_{GT}\text{[G[i],T[i]]} + 
\beta_{GI}\text{[G[i],I[i]]} + 
\beta_S\text{[S[i]]}$$

$\beta_0$ is the overall average OCR per 1000 cells in all wells and across injections. 
As before, we used $\text{Lognormal(0.54, 0.32)}$ as the prior distribution. 
$\beta_G$, $\beta_T$ and $\beta_I$ are OCR deflections due to group (healthy or patient), treatment (glucose or galactose), and injections, respectively. 
To take into account the fact that cell lines derived from the same subject were used for all levels of treatments and injections, $\beta_S$ is used to estimate the deflection due to subject factor. 
The interaction term, $\beta_{IT}$ denotes whether deflection due to injections is dependent on treatment factor. 
Similarly, the other two interaction terms, $\beta_{GT}$ and $\beta_{GI}$ denote whether deflection caused by treatment and injection factor are dependent on group factor.

>>>>

---
output: word_document
bibliography: /home/xiang/Documents/Bibliography/seahorse.bib
csl: plos-one.csl
---

OCRbayes: A Bayesian hierarchical modeling framework for Seahorse extracellular flux oxygen consumption rate data analysis

@@

Mitochondrial dysfunction is involved in many complex diseases. 
Efficient and accurate evaluation of mitochondrial functionality is crucial for understanding pathology as well as facilitating novel therapeutic developments. 
As a popular platform, Seahorse extracellular flux (XF) analyzer is widely used for measuring mitochondrial oxygen consumption rate (OCR) in living cells. 
A hidden feature of Seahorse XF OCR data is that it has a complex data structure, caused by nesting and crossing between measurement cycles, wells and plates. 
Surprisingly, statistical analysis of Seahorse XF data has not received sufficient attention, and current methods completely ignore the complex data structure, impairing the robustness of statistical inference. 
To rigorously incorporate the complex structure into data analysis, here we developed a Bayesian hierarchical modeling framework, OCRbayes, and demonstrated its applicability based on analysis of published data sets. 
We showed that OCRbayes can analyze seahorse XF OCR data with either single or multiple plates, and it has potential to be used for diagnosing patients with mitochondrial diseases.

@@

Mitochondria produce energy primarily by transferring electrons along the electron transport chain (ETC) on the inner membrane. 
Along the ETC, there are four complexes (including complex I, II, III and IV), building up a proton gradient that is ultimately used by ATP synthase to generate ATP. 
Since electrons are primarily accepted by O~2~ to produce H~2~O, the activity of ETC can be assessed by measuring oxygen consumption rate (OCR). 
As a reliable and efficient platform, Seahorse XF analyzer is a multiwell plate based respirometry assay that is widely used to quantify OCR in living cells [@Wu2007]. 
Typically, the seahorse XF analyzer measures the OCR of cells in a 96-well plate under different ETC perturbation scenarios, used for assessing mitochondrial functionality such as maximal respiration, leak respiration and ATP-linked respiration [@@Chacko2014]. 
Seahorse XF Cell Mito Stress Test consists of four phases including an initial phase, followed by three consecutive injections, including oligomycin, FCCP (carbonyl cyanide-4-(trifluoromethoxy)phenylhydrazone) and a mixture of rotenone and antimycin A. 
In the initial phase, no compound is injected and ETC is not perturbed. 
Once oligomycin is injected in the assay, proton translocation through ATP synthase is blocked, leading to a reduction in OCR. 
Next, injection of FCCP allows protons to move into the mitochondrial matrix independent of the ATP synthase, pushing the oxygen consumption to the maximum level. 
In the end, a mixture of rotenone and antimycin A is injected to inhibit complex I and III, shutting down mitochondrial respiration.

A standard seahorse run includes three measurement cycles for each phase. 
Every measurement cycle starts by lowering the cartridge and creating a temporary semi-closed 7 $\mu$L chamber [@Gerencser2009]. 
During a measurement cycle, fluorescent oxygen sensors capture oxygen concentration changes in the chamber and outputs OCR. 
The Seahorse XF analyzer measures OCR in tens to hundreds of thousands of cells per well and typically requires 4–5 replicate wells per experimental group [@Artyomov2020].
A single 96-well plate can contain multiple experimental groups. 
Since measurement cycles are nested within phases, phases are crossed with wells, and wells are nested in experimental groups, a very complex structure is inherently embedded in the Seahorse XF OCR data. 
Surprisingly, current Seahorse XF data analyses most often ignore this complex structure, and by default, data sets are often chopped into subgroups followed by performing ANOVA-like statistical tests. 
Although there are tools developed for seahorse data analysis, none of them takes the complexity of the data structure into account. 
In addition, these methods differ in how to extract OCR value for each phase. 
For instance, OCR-Stats [@Yepez2018] took OCR value from the last measurement cycle in phase 1, the minimal OCR value in phase 2, the maximal OCR value in phase 3 and the minimal OCR value in phase 4. In contrast, SHORE [@Nicholas2017] took the mean OCR value in phase 1, the median OCR value in phase 2, the maximal OCR value in phase 3, and the median OCR value in phase 4.

A natural way to incorporate the complex structure into data analysis is to use hierarchical modeling, which has been extensively developed for analyzing gene expression data  [@Robinson2010a;@Love2014a]. 
Here we developed a Bayesian hierarchical modeling framework, OCRbayes, for the Seahorse XF OCR data analysis. 
OCRbayes is based on a fully Bayesian approach due to its flexibility and convenience for constructing the hierarchical models. 
We analyzed a published OCR data set OCR data set to demonstrate the applicability of our approach analyzing Seahorse XF OCR data as well as its potential usage for mitochondrial disease diagnostics.

@@

To demonstrate how OCRbayes can be used for Seahorse XF OCR data analysis, we analyzed 26 human fibroblasts that were derived from patients that were suspected to have mitochondrial dysfunction. 
In particular, six of these cell lines were derived from patients with rare mutations in genes associated with an established cellular respiratory defect, and they serve as positive controls. 
Meanwhile, there are two cell lines that showed no respiratory defects in earlier studies [@Powell2015;@Kremer2016] and they served as negative controls. 
Since fold change of maximal respiration is the most common metric used for identifying mitochondrial dysfunction, we compared the maximal respiration of all these 26 cell lines against the control cell line.

## Data

We downloaded the seahorse XF OCR data presented in OCR-Stats [@Yepez2018] from the journal website https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0199938#sec022. 
This data set contains 203 human fibroblasts assayed in 126 plates. 
The normal human dermal fibroblasts (NHDF) is used as control. 
The rest 202 cell lines were derived from patients suspected to suffer from rare mitochondrial diseases. 
Among these 202 cell lines, 26 fibroblast cell lines were grown independently and were measured in multiple plates. 
The rest 176 fibroblast cell lines were measured in a single plate. 
We then spitted the data set into two subsets. 
One subset, which contains the 176 fibroblasts that were assayed in a single plate, was used for estimating prior distributions. 
The other subset, which contains the 26 fibroblasts, was used for benchmarking.

Since dealing missing observations is not the focus of this study, we removed the entire measurement values of a well if it contains a missing OCR value. 
In addition, since we focused on the standard mitochondrial stress test assay, we removed any well if that well does not have 4 phases with 3 measurement cycles each. 
For the prior distribution subset, we had 177 (including the control cell line NHDF) cell lines on 78 plates after filtering. 
Since NHDF cell line was present in all 78 plates, the OCR values derived from NHDF cell line were used for estimating the prior distribution of between plate variation.

@@

The Bayesian hierarchical models were implemented in Stan (version 2.19.3) [@Carpenter2017]. 
We fitted the model by running Hamiltonian Markov Chain Monte Carlo. 
We ran four Markov chains with 2000 iterations in each chain. 
The code can be found at https://github.com/XiangZhangSC/seahorse. 
We also provided the R code of OCRbayes in the supplementary file.

@@

$$\text{Basal respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, rotenone} \text{ (Equation 1)}$$
$$\text{ATP-linked respiration} = \text{OCR}_\text{per 1k cells, initial} - \text{OCR}_\text{per 1k cells, oligomycin} \text{ (Equation 2)}$$
$$\text{Proton leak} = \text{OCR}_\text{per 1k cells, oligomycin} - \text{OCR}_\text{per 1k cells, rotenone} \text{ (Equation 3)}$$
$$\text{Spare respiratory capacity} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, initial} \text{ (Equation 4)}$$
$$\text{Maximal respiration} = \text{OCR}_\text{per 1k cells, FCCP} - \text{OCR}_\text{per 1k cells, rotenone} \text{ (Equation 5)}$$

## Statistical analysis

For each cell line study, we compared the patient cell line to the control cell line (NHDF). 
Since maximal respiration is the primary readout of mitochondrial defect in literature, and was also evaluated in the OCR-Stats, here we reported the mean log2 fold change of maximal respiration together with the False Discovery Rate (FDR). 
FDR was calculated based on the posterior error probability ($1 - P(log2(\frac{\text{patient}}{\text{control}}|\text{data}) < 0)$), where $P$ represents the posterior distribution after fitting the Bayesian model to the experimental OCR data. 
If the FDR below 0.05, we consider that difference between patient and control cell line was statistically significant.

@@

## OCRbayes: A Bayesian hierarchical modeling framework for Seahorse OCR data analysis

In order to incorporate the complex data structure into the analysis of seahorse XF OCR measurements, we developed a Bayesian hierarchical modeling framework, OCRbayes. 
In this study, we focused on experimental data containing two groups, patient and control. 
The mathematical description of our Bayesian hierarchical modeling framework is shown below:

$$\text{OCR}_\text{obs}\text{[i]} \sim \text{Lognormal(logOCR}_\text{true}\text{[P[i],W[i],I[i]]}, \text{logOCR}_\text{se}\text{[P[i],I[i]])} \text{ (Equation 6)}$$
$$\text{logOCR}_\text{true}\text{[P[i],W[i],I[i]]} \sim \text{Normal(}\text{OCR}_\text{per 1k cells}\text{[P[i],G[i],I[i]]} \times \text{N}_\text{cells}\text{[W[i]]}, \sigma_\text{well}\text{[P[i],G[i],I[i]])} \text{ (Equation 7)}$$
$$\text{OCR}_\text{per 1k cells}\text{[P[i],G[i],I[i]]} \sim \text{Lognormal(}\mu_{\text{OCR}_\text{per 1k cells}}\text{[G[i], I[i]]}, \sigma_\text{plate}\text{[G[i], I[i]])} \text{ (Equation 8)}$$

From our perspective, there are three sources of technical variations including 1) between measurement cycle variation, 2) between well variation and 3) between plate variation in the Seahorse XF OCR data.

The between measurement cycle variation refers to the variability between the three OCR values generated by the three measurement cycles within an injection phase. 
In Equation 6, we used a lognormal distribution with the true OCR value at the log scale ($\text{logOCR}_\text{true}$) and the between measurement cycle standard deviation (also at the log scale, $\text{logOCR}_\text{se}$) to model each observed OCR value ($\text{OCR}_\text{obs}$). 
$\text{OCR}_\text{obs}$ is a vector of length $\text{N}_\text{plates} \times \text{N}_\text{wells} \times \text{N}_\text{injections} \times \text{N}_\text{Measurement}$, where $\text{N}_\text{plates}$, $\text{N}_\text{wells}$, $\text{N}_\text{injections}$ and $\text{N}_\text{Measurement}$ are number of plates, wells, injections and measurement cycles, respectively. 
The $\text{logOCR}_\text{true}$ refers to $\text{N}_\text{plates}$ layers of matrices. 
Each layer is a matrix with $\text{N}_\text{wells}$ rows and $\text{N}_\text{injections}$ columns. 
$\text{logOCR}_\text{se}$ is a matrix of $\text{N}_\text{plates}$ rows and $\text{N}_\text{injections}$ columns.

In a typical seahorse XF assay, the same cell line from the same experimental group is often seeded in multiple wells. 
However the number of cells in these replicate wells are not identical, thus causing variation in OCR values between these replicate wells. 
To adjust for the effect of cell number difference, in Equation 7 we modeled the true OCR value as a function of the cell number, and from there estimated $\text{OCR}_\text{per 1k cells}$, which represents OCR value per 1000 cells that received the same treatment (or from the same group) and injection.
$\text{OCR}_\text{per 1k cells}$ is a three dimensional matrix with $\text{N}_\text{Groups}$ layers, and each layer has $\text{N}_\text{Groups}$ rows and $\text{N}_\text{injections}$ columns. 
$\text{N}_\text{cells}$ is a vector of length $\text{N}_\text{wells}$, and every entry represents the cell number in that well. 
In addition to cell number difference, initial conditions, treatment concentration, or fluorophore sleeve calibration can also contribute to between well variation. 
Since these factors were difficult to be measured experimentally, we can only estimate an overall effect based on the residual parameter $\sigma_\text{well}$. 
Thus, $\sigma_\text{well}$ represents the well-to-well variation after accounting for cell number difference.

Last but not the least, the seahorse XF assay for a patient cell line is often repeated on different days and on more than one plates. 
Due to differences in temperature, seeding time, growth time, growth medium, sensor cartridge as well as treatment efficiency, the OCR outcomes will differ between plates. 
To take into account the between plate variation, in Equation 8 we used another lognormal distribution with the true OCR value per 1000 cells at log scale ($\mu_{\text{OCR}_\text{per 1k cells}}$) and the between plate standard deviation ($\sigma_\text{plate}$) to model the OCR value per 1000 cells $\text{OCR}_\text{per 1k cells}$.

In this study, since our case studies focused on human fibroblast cells, we used informative prior distributions for $\text{logOCR}_\text{se}$, $\sigma_\text{well}$, $\sigma_\text{plate}$ and $\mu_{\text{OCR}_\text{per 1k cells}}$. 
Since values of all these parameters must be positive, we used four lognormal distributions. 
We estimated the lognormal distribution parameters based on the Seahorse XF OCR data in OCR-stats [@Yepez2018] by running the fitdistr function built in the MASS r package. 
In this study, our prior distributions are
$$
\begin{align}
\text{logOCR}_\text{se} & \sim \text{Lognormal(-3.23, 0.79)} \\
\sigma_\text{well} & \sim \text{Lognormal(-1.62, 0.53)} \\
\sigma_\text{plate} & \sim \text{Lognormal(-1.18, 0.05)} \\
\mu_{\text{OCR}_\text{per 1k cells}} & \sim \text{Lognormal(0.3, 0.79)}
\end{align}
$$

## OCRbayes uses OCR measurements as inputs and outputs posterior distributions for respiration metrics

OCRbayes works with the raw OCR data, and estimates the OCR per 1000 cells during initial phase (Int1), oligomycin phase (Int2), FCCP phase (Int3) and rotenone phase (Int4) (Figure 1A and 1B). 
Afterwards, the posterior OCR per 1000 cells values were transformed into various respiration metrics, such as ATP-linked respiration, basal respiration, maximal respiration, proton leak and spare respiratory capacity (Figure 1C). 
As an example, here we showed patient 65126 and the control cell line (NHDF) on the corresponding plates. 
The posterior distributions showed the range of values for a respiration parameter. 
The spread of the posterior distributions represented the uncertainty, which is resulted from technical variations that will be discussed in the next section.

![Figure 1 (A) Raw OCR values for patient 65126 and control cell line in two plates. (B) Posterior distributions for OCR per 1000 cells in patient 65126 and control cell line during initial phase (Int1), oligomycin phase (Int2), FCCP phase (Int3) and rotenone phase (Int4). (C) Posterior distributions for respiration metrics in patient 65126 and control cell line.](./posterior_respiration_examples.jpeg)

## OCRbayes accounts for various technical variations during seahorse XF OCR data analysis

To visualize the technical variations, we used the cell line 76065 as an example. 
This cell line was measured on five different plates, allowing us to visualize all three technical variations.  
We observed that in the initial phase (Int1) and FCCP phase (Int3), the between plate variation is dominant source of technical variation, followed by between well and between measurement cycle variations. 
However, in the oligomycin phase (Int2) and rotenone phase (Int4), the between well variation can be as important as the between plate variation.

![Figure 2 Technical variations in seahorse XF OCR data of experiments based on cell line 76065 and control cell line. Int1, Int2, Int3 and Int4 represent initial, oligomycin, FCCP and rotenone phase. All the variations were on log scale.](./technical_variations_76065.jpeg)

## Benchmark

To demonstrate that OCRbayes works properly, we applied it to analyze a published seahorse XF OCR data set containing 26 patient cell lines as well as a control cell line. 
In a previous study, 6 cell lines were labeled as positive controls and 2 cell lines were labeled as negative controls. 
Based on our analysis, we found six patient cell lines that had lower maximal respiration compared to the control cell line with False Discovery Rate (FDR) below 0.05. 
Five of these six cell lines were actually the positive controls. 
Meanwhile, the two negative controls had FDR above 0.05.

![Figure 3 Maximal respiration change patient vs. control on multiple plates. (A) average log2 fold change (y-axis) of maximal respiration of all cell lines repeated across plates (x-axis) and their respective controls, sorted by the False Discovery Rate (FDR). Red and blue dots represent positive and negative controls, respectively. (B) similar to (A), but depicting FDR. Red dashed line represents FDR = 0.05.](./benchmark.jpeg)

Our analysis successfully recalled  all negative controls and five out of six positive controls. 
Interestingly, our analysis showed that there was no significant reduction of maximal respiration in patient 61818, which was used as a positive control. 
This patient cell line was measured by seahorse XF OCR assay on three independent plates on three different days. 
We analyzed these three repeated experiments separately as well as combined them together. 
We found that only the second experiment showed difference in maximal respiration.

![Figure 4 (A) Raw OCR values for patient cell line 61818 in three repeated plates. (B) Posterior distributions of maximal respiration of patient 61818 and the control cell line in the same plate. The top three panels represent analysis of three repeated experiments separately. The last panel represents the data analysis outcome after combing all three repeated experiments.](./posterior_maximal_respiration_61818.jpeg)

## OCRbayes can facilitate decision making about whether testing on multiple plates is necessary

A frequent question regarding seahorse XF analysis is that whether it is enough to perform the assay on a single plate or do we really need repeat the assay multiple times on different plates. 
To address this question, we used the two positive control cell lines 67375 and 76065 as examples, since in our analysis they were shown to have the maximal and minimal effect size among the positive controls, respectively.

The cell line 67375 was measured on two plates. 
Based on the first assay, our analysis shewed that the posterior probability of patient 67375 having decreasing maximal respiration than the control was 93.0%. 
By repeating the experiment once more, the posterior probability increased to 99.4%. 
In contrast, the cell line 76065 was measured on five different plates. 
Our analysis showed that based on the first assay, the posterior probability of patient 76065 having lower maximal respiration than the control was 40.6%. 
Repeating the assay once increased the posterior probability from 40.6% to 72.0%. 
When the assay was repeated for the third, fourth and fifth time, the corresponding posterior probabilities increased to 78.8%, 87.9% and 89.9%, respectively.

![Figure 5 (A) Posterior probability of decreasing maximal respiration in patient 67375 compared to the control cell line in the scenario with 1 or 2 plates. (B) Posterior probability of decreasing maximal respiration in patient 76065 compared to the control cell line in the scenario with 1, 2, 3, 4 or 5 plates](./posterior_max_resp_67375_76065.jpeg)
